---
title: "EDS241: Take Home Final"
author: "Scout Leonard"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: no
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)


# load packages
packages=c("stargazer",
           "here",
           "tidyr",
           "dplyr",
           "stringr",
           "janitor", 
           "cowplot",
           "ggplot2",
           "tinytex",
           "datasets",
           "tibble",
           "estimatr",
           "knitr",
           "plm",
           "car")

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

#devtools::install_github('rstudio/rmarkdown')
options(scipen=999) # not scientific notation
```

# Read in and clean data

```{r}
#read in data 
housing <- read.csv(here("data", "KM_EDS241.csv")) %>% 
  mutate(nearinc = as.factor(nearinc),
         year = as.factor(year))
```

\newpage

# Question A

**Using the data for 1981, estimate a simple OLS regression of real house values on the indicator for being located near the incinerator in 1981. What is the house value “penalty” for houses located near the incinerator? Does this estimated coefficient correspond to the ‘causal’ effect of the incinerator (and the negative amenities that come with it) on housing values? Explain why or why not.**

\noindent The code chunk below estimates a simple OLS regression of real house values on the indicator for being located near the incinerator in 1981.

```{r}
#filter housing for only observations from 1981
housing_1981 <- housing %>% 
  filter(year == 1981)

#model simple ols regression
ols_mod_1981 <- lm_robust(rprice ~ nearinc, 
                          data = housing_1981)
```

\noindent The code chuck below inputs the outputs of the simple OLS regression `ols_mod_1981` into a table using the function `kable()`. 

```{r}
#create table with regression results
ols_mod_1981_table <- tidy(ols_mod_1981)

#print table 
ols_mod_1981_table %>% 
  select(term, estimate, std.error, p.value, conf.low, conf.high) %>%
  kable()
```

\noindent The house value "penalty" for houses located near the incinerator is `r abs(round(ols_mod_1981$coefficients[2], digits = 2))` dollars. 

\noindent This estimated coefficient does respond to the 'causal' effect of the incinerator, although there can be other variables which effect the penalty for houses located near the incinerator, due to omitted variables bias. 

\newpage

# Question B

**Using the data for 1978, provide some evidence the location choice of the incinerator was not “random”, but rather selected on the basis of house values and characteristics. [Hint: in the 1978 sample, are house values and characteristics balanced by nearinc status?]**

```{r}
#filter housing for only observations from 1981
housing_1978 <- housing %>%
  filter(year == 1978)
```

## Houses Near the Incinerator: 1978 Summary Exploration

\noindent The code chunk below filters the data for observations from the housing in 1978 subset for only the houses near the incinerator. It returns the summary statistics for each of the seven variables for this subset of a subset.

```{r}
housing_1978_nearinc <- housing_1978 %>% 
  filter(nearinc == 1)

summary(housing_1978_nearinc) %>% 
  kable()
```

\newpage

## Houses Far from the Incinerator: 1978 Summary Exploration

\noindent The code chunk below filters the data for observations from the housing in 1978 subset for only the houses far from the incinerator. It returns the summary statistics for each of the seven variables for this subset of a subset.

```{r}
housing_1978_farinc <- housing_1978 %>% 
  filter(nearinc == 0)

summary(housing_1978_farinc) %>% 
  kable()
```

\noindent In comparing summary statistics for variables from 1978 Andover houses from near and far from the incinerator, we can see that houses far from the incinerator have higher mean and median area, land, and price, and lower mean and median age. This suggests location choice of the incinerator _may not_ be random. 

\noindent In the code chunk below, linear regressions return the average difference in outcomes between houses near and far from the incinerator. 

## Simple OLS: Average Differences

### Average Difference in Area of the House

```{r}
mod1_1978 <- lm_robust(area ~ nearinc, data  = housing_1978)

#create table with regression results
mod1_1978_table <- tidy(mod1_1978)

#print table 
mod1_1978_table %>% 
  select(term, estimate, std.error, p.value, conf.low, conf.high) %>%
  kable()
```

### Average Difference in Area of the Lot

```{r}
mod2_1978 <- lm_robust(land ~ nearinc, data = housing_1978)

#create table with regression results
mod2_1978_table <- tidy(mod2_1978)

#print table 
mod2_1978_table %>% 
  select(term, estimate, std.error, p.value, conf.low, conf.high) %>%
  kable()
```

\noindent The results indicate that on average, houses near the incinerator have an area `r abs(round(mod1_1978$coefficient[2], digits = 2))` square feet smaller than houses far from the incinerator. 

\noindent On average, houses near the incinerator have `r abs(round(mod2_1978$coefficient[2], digits = 2))` fewer square feet of lot compared to houses far from the incinerator. 

\noindent These differences show that on average, houses near and far from the incinerator have statistically different characteristics. This serves as evidence that the siting of the generator was not "random." 

\newpage

# Question C

**Based on the observed differences in (b), explain why the estimate in (a) is likely to be biased downward (i.e., overstate the negative effect of the incinerator on housing values).**

\noindent Based on the observed differences in (b), the estimate in (a) is likely to be biased downward (overstate the negative effect of the incinerator on housing values), because variables such as size and area probably already brought down the value of houses where the incinerator was placed, but the effect of the incinerator on housing values emphasized the difference in values between houses near and far from the incinerator in North Andover. 

\newpage

# Question D

**Use a difference-in-differences (DD) estimator to estimate the causal effect of the incinerator on housing values without controlling for house and lot characteristics. Interpret the magnitude and sign of the estimated DD coefficient.**

```{r}
dd <- lm_robust(formula = rprice ~ nearinc +
           year,
         data = housing)

summary(dd)
```

The DD value tells us that by holding time constant, the houses near the incinerator cost `r abs(round(dd$coefficient[2], digits = 2))` less than houses far from the incinerator. 

\noindent

\newpage

# Question E

**Report the 95% confidence interval for the estimate of the causal effect on the incinerator in (d).**

```{r}
dd_ci_95 <- confint(dd)

dd_ci_95
```

The 95% confidence interval for the estimate of the causal effect on the incinerator is (d) is `r round(dd_ci_95[2], digits = 2)` to `r round(dd_ci_95[2,2], digits = 2)`

\newpage

# Question F

**How does your answer in (d) changes when you control for house and lot characteristics? Test the hypothesis that the coefficients on the house and lot characteristics are all jointly equal to 0.**

```{r}
dd_2 <- lm_robust(formula = rprice ~ nearinc +
             year +
             age +
             rooms +
             area +
             land,
         data = housing)

summary(dd_2)
```

\newpage

When you control for house and lot characteristics, the difference in price between houses near and far from the incinerator, holding all else equal is, is smaller than in the model in which we do not control for these characteristics. This smaller Beta 1 value, however, is no longer statistically significant in the new model. This means that beings near or far from the incinerator may not be a good indicator for the price of a house in North Andover.

```{r}
#linear hypothesis
linearHypothesis(model = dd_2, 
                 c("area = 0", 
                   "land = 0", 
                   "rooms = 0", 
                   "age = 0"), 
                   white.adjust  = "hc2")
```

With this P value, I would reject the null hypothesis that the coefficients on the house and lot characteristics are all jointly equal to 0. 

\newpage

# Question G

**Using the results from the DD regression in (f), calculate by how much did real housing values change (for the control group) on average between 1978 and 1981.**

Between 1978 and 1981, real housing values increased by `r round(dd_2$coefficient[3], digits = 2)` dollars if all other variables are held equal. 

\newpage

# Question H

**Explain (in words) what is the key assumption underlying the causal interpretation of the DD estimator in the context of the incinerator construction in North Andover.**

A key assumption of the DD estimator corresponds to the parallel trend assumption, where the control group, the houses far from the incinerator, provide a valid counterfactual for the temporal evolution of mean outcomes (housing prices) in the treatment group, houses near the incinerator, in the absence of a change in treatment (placement of the incinerator).
